Statistically-estimated tree composition for the northeastern United States \\ at the time of Euro-American settlement

### 2016-02-03 paperPlos_track.tex
revised and added for plos revision january 2016

In \emph{Data} we describe the data sources, while in \emph{Statistical model} we describe 
the statistical model used to create the data product. In In \emph{Model comparison}
 we quantitatively compare competing
statistical specifications, and in \emph{Data Product}
we describe the final data product. In \emph{Discussion}
we discuss the uncertainties estimated by and the
limitations of the statistical model, and we list related data products
under development.

@@

The raw data were obtained from land division survey records collated
and digitized from across the northeastern U.S. by a number of researchers
(Fig. \ref{fig:Spatial-domain}). For the states of Minnesota, Wisconsin,
Illinois, Indiana, and Michigan (the midwestern subdomain), digitized
data are available at PLS survey point locations and have been aggregated
to a regular 8 km grid in the Albers projection. (Note that for Indiana
and Illinois, at the moment trees are associated with township centroids
and then assigned to 8 km grid cells based on the centroid, but in
the near future we will have point locations available for each tree.)
For the states of Ohio, Pennsylvania, New Jersey, New York and the
six New England states (the eastern subdomain), data are aggregated
at the township level. We make predictions for all
of the states listed above; these constitute our core domain. There
are also data from a single township in Quebec and a single township
in northern Delaware; these data help inform predictions
in nearby locations within our core domain, but predictions are not
made for Quebec or Delaware. Digitization of PLS data in Minnesota,
Wisconsin and Michigan is essentially complete, with PLS data for
nearly all 8 km grid cells, but data in Illinois and Indiana represent
a sample of the full set of grid cells, with survey record transcription
ongoing. Data for the eastern states are available for a subset of
the full set of townships covering the domain; the TPS data for some
townships were lost, incomplete, or have not been located \cite{Cogb:etal:2002}.

@@

Extensive details on the upper Midwest (Minnesota, Wisconsin, Michigan)
data and processing steps are available \cite{goring2015composition};
key elements include the use of only corner points, the use of only
the two closest trees at each corner point, spatially-varying correction
factors for sampling effort, and a standardized taxonomy table. The
lower Midwest (Illinois, Indiana) data were purchased from the Indiana
State Archives (Indiana) and Hubtack Document Resources (hubtack.com;
Illinois) and processed using similar steps as for the upper Midwest
data. Digitization of the Illinois and Indiana data is still underway,
so many grid cells contained no data at the time the statistical model
was fit. Note that the number of trees per grid cell varies depending
on the number of survey points in a cell, with an average of 124 trees
per cell. The gridded data at the 8 km resolution for the midwest
subdomain are available through the NIS data portal as
DOI:10.5072/FK2/e9c144ab83af70e29b8d6647aaae58f5 \cite{Gori:etal:data:2016}.
The TPS data were compiled by C.V. Cogbill from a myriad of archival
sources representing land division surveys conducted in connection
with local settlement and are available through the
NIS data portal as DOI:10.5072/FK2/4d44aebe71dbf65290d8ce1bde03598b
and DOI:10.5072/FK2/88e167487b22d5f657e7ce7e033772e1 \cite{Cogb:dataNE:2016,Cogb:dataOH:2016}.

@@

Our domain is a rectangle covering all of the states using a metric
Albers (Great Lakes and St. Lawrence) projection (PROJ4: EPSG:3175),
with the rectangle split into 8 km cells, arranged in a 296 by 180
grid of cells, with the centroid of the cell in the southwest corner
located at (-71000 m, 58000 m). For the midwestern subdomain we use
the western-most 146 by 180 grid of cells when fitting
the statistical models. For the eastern subdomain we use the eastern-most
180 by 180 grid of cells and then omit 23 rows of cells in the north
and 17 rows of cells in the south, as these grid cells are outside
of the states containing data.

@@

We fit a Bayesian statistical model to the data, with two primary
goals:
\begin{enumerate}
\item To estimate composition on a regular grid across the entire domain,
filling gaps where no data are available, and
\item To quantify uncertainty in composition at all locations. Even in grid
cells and townships with data, we wish to quantify uncertainty because
the empirical proportions represent estimates of the true proportions
that could be calculated using the full population of all the trees
in a grid cell or township.
\end{enumerate}
At a high level, the Bayesian statistical model estimates
composition across the domain, even in locations with sparse or no
data, by combining the raw composition data with the assumption that
composition varies in a smooth spatial fashion across the domain.
The information in the data is quantified by the data model, also
known as the likelihood. The assumption of smoothness is built into
the model by representing the true unknown spatially-varying composition
using a statistical spatial process representation that induces smoothing
of estimates across nearby locations. This spatial process representation
is a form of prior distribution and is a function of model parameters
called hyperparameters that determine the correlation structure of
the process and are also estimated based on the data.

@@

In the remainder of this section we provide the technical
specification of the model and of the computations involved in fitting
the model.

@@

The proportions, $\theta_{p}(s_{i}),\, p=1,\ldots,P$, are modeled
spatially by a set of \emph{P} Gaussian spatial processes, one per taxon,
$\alpha_{p}(s_{i}),\, p=1,\ldots,P$. This collection of processes
defines a multivariate spatial process for composition. The $\alpha_{p}(s)$
processes are defined on the 8 km grid, $\alpha_{p}=\{\alpha_{p}(s_{1}),\ldots,\alpha_{p}(s_{m})\}$
for the \emph{m} grid cells. In  \emph{Latent variable model}
we introduce a multinomial probit model that relates the $\alpha_{p}(s)$
processes to the proportion processes, $\theta_{p}(s)$, via the introduction
of latent variables, with an implicit sum-to-one constraint, $\sum_{p=1}^{P}\theta_{p}(s)=1$.
A multinomial probit model is similar to logistic
regression, used for modeling a binary outcome based on an underlying
probability the outcome will occur, but generalizes to modeling a
categorical variable based on probabilities for each category.

The critical component of the statistical model is the representation
of $\alpha_{p}(s)$ as a spatial process. A spatial
process is a statistical representation that models spatially-correlated
values. It provides a prior structure that serves to smooth across
noise in the observations and allows for prediction at locations based
on information from nearby locations, including interpolation to locations
with no data. Apart from the sum-to-one constraint, the taxa are considered
to be independent in the prior. We did not want to impose any structure
that ties the different taxa together, as any correlation will likely
vary across space.

In the next section, we consider two spatial models to define the
structure of the $\alpha_{p}(s)$ processes, a standard conditional
autoregressive model \cite{Bane:etal:2004} and a Gaussian Markov
random field (MRF) approximation to a Gaussian process with \matern
covariance \cite{Lind:etal:2011}. These models are
specific statistical formulations of spatial processes that represent
spatial correlation by defining neighborhoods around each location
that are used to help inform predictions at the location.

@@

MRF models represent the neighborhood information
by working directly with the precision matrix (the
inverse of the covariance matrix) of the values of the spatial process,
so calculation of the prior density of $\alpha_{p}$ is computationally
simple \cite{Rue:Held:2005}. However, in situations
in which the likelihood is not normal, such
as our multinomial likelihood, it can be difficult to set up effective
MCMC algorithms that are able to move in the high-dimensional space
of $\alpha_{p}$. The latent variable representation
helps to alleviate this problem. Next we describe two alternative
spatial models that we considered; in \emph{Model comparison},
we evaluate the models on held-out data to choose
between the two.

@@

Our first model is a standard conditional autoregressive (CAR) model;
technical details can be found in \cite{Bane:etal:2004}. We use
a standard form of this model that treats the four
cardinal neighbors of each grid cell as the neighbors of the grid
cell. The corresponding precision matrix has diagonal elements, $Q_{ii}$,
equal to the number of neighbors for the \emph{i}th area (i.e., four except
for cells on the boundary of the domain), while $Q_{ik}=-1$ (the
negative of a weight of one) when areas \emph{i} and \emph{k} are neighbors
and $Q_{ik}=0$ when they are not. This gives the following model
for the values of $\alpha_{p}(s_{i})$ collected as a vector across
all of the grid cells, $i=1,\ldots,m$: 
\[
\alpha_{p}\sim\mbox{N}(0,\sigma_{p}^{2}Q^{-}).
\]
The use of the generalized inverse notation indicates that \emph{Q} is
not full-rank, but is of rank $m-1$; this gives an improper prior
on an implicit overall mean for the process values. Note
that we specify an explicit mean of zero because a non-zero mean would
not be identifiable in light of the implicit mean. This specification
is called an \textit{intrinsic conditional autoregression (ICAR)}
and we can write $Q=D-C$ where \emph{C} is the $m\times m$ adjacency
matrix defining the neighborhood relation of the locations; that is
$C{}_{ik}=1$ if locations \emph{i} and \emph{k} are neighbors and zero otherwise.
The matrix $D$ is an $m\times m$ diagonal matrix containing the
row sums of matrix $C$ as the diagonal entries, $D{}_{ii}={\displaystyle \sum_{k=1}^{m}C{}_{ik}}.$

@@

Gaussian processes (GP) are also standard models for spatial processes \cite{Bane:etal:2004}.
GP models are computationally challenging for large datasets because
of computational manipulations involving large covariance
matrices. Given this, \cite{Lind:etal:2011} proposed a new framework
for using Gaussian MRFs (GMRFs) as approximations to GPs, based on
the use of stochastic partial differential equations (SPDEs).

Gaussian processes are generally constructed using
one of a number of correlation functions that define how the strength
of correlation between the values of the process at two locations
decays as a function of the distance between the locations. We consider
Gaussian processes in the commonly-used \matern
class, using the following parameterization of the \matern correlation
function, 
\begin{equation}
R(d)=\frac{1}{\Gamma(\nu)2^{\nu-1}}\left(\frac{2\sqrt{\nu}d}{\rho}\right)^{\nu}\mathcal{K}_{\nu}\left(\frac{2\sqrt{\nu}d}{\rho}\right),\label{eq:Matern}
\end{equation}
where $d$ is Euclidean distance, $\rho$ is the spatial range parameter,
and $\mathcal{K}_{\nu}(\cdot)$ is the modified Bessel function of
the second kind, whose order is the smoothness (differentiability)
parameter, $\nu>0$. $\nu=0.5$ gives the exponential covariance.
For any pair of locations, $R(d)$ defines the correlation of the
process, (i.e., $\alpha_{p}(s)$ in our context), as a function of
the distance between the locations. Considering all pairs of locations,
this defines a correlation matrix for all locations of interest.

The approach of \cite{Lind:etal:2011} allows us to consider MRF
approximations to the \matern-based GP for $\nu=1$ and $\nu=2$.
Our second spatial model is this Lindgren approximation for \matern-based
GPs with $\nu=1$. To implement this Lindgren model, one modifies
the \emph{Q} matrix defined previously as follows based
on the technical specification of the precision matrix provided in
\cite{Lind:etal:2011}. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
elements of \emph{Q} are $4+a^{2}$. The entries corresponding to cardinal
neighbors are $-2a$. Those for diagonal neighbors are $2$, and those
for 2nd-order cardinal neighbors are $1$. This extends the neighborhood
structure relative to the CAR model and parameterizes it as a function
of $\rho$.

The primary difference between the CAR and Lindgren models is that
the Lindgren model provides an additional degree of freedom by estimating
$\rho$. In particular $\rho$ allows us to estimate the locality
of the spatial smoothing. As $\rho$ decreases, the
model uses increasingly localized data to estimate the compositional
proportions at a given location, effectively averaging the empirical
proportions over smaller neighborhoods. In general, the \cite{Lind:etal:2011}
model will generally provide for a smoother estimate than the CAR
model \cite{Paci:2013}.

@@

\noindent The ICAR specification contains a set of hyperparameters
$\{\sigma_{p}^{2}\},$ $p=1,\ldots P$. Following \cite{Gelm:2006}
we use a uniform distribution on each $\sigma_{p}$ parameter, with
upper bound of 1000. For the SPDE model we also have hyperparameters
$\{\mu_{p}\}$, which we give flat, non-informative priors (truncated
at $\pm10$), and $\{\rho_{p}\}$, which we give uniform priors on
the interval $(0.1,\exp(5))$. These various hyperparameters
are unknown parameters that control the spatial structure of the two
spatial models and are estimated from both the data and the prior
distributions just specified based on the Bayesian approach.

@@

We compared the CAR and SPDE models by holding out
data from the fitting process and assessing the fit of the model
on the held-out data. We used two experiments with
held-out data:
\begin{enumerate}
\item The first experiment used a subregion containing most of Minnesota
and a small amount of western Wisconsin, defined to be the cells whose
x-coordinate was less than 300,000 m (this defines a north-south line
that approximately goes through Duluth, Minnesota) and hereafter referred
to as the ``Minnesota subregion''. We chose this subregion for evaluation
because of its high data density, allowing us to experiment with the
effects of increasing data sparsity on model performance. We held
out all data from 95\% of the cells in this Minnesota subregion, with
cells selected at random. This was meant to assess the ability of
the model to interpolate from a sparse set of cells/townships and
mimics the limited data in Illinois and Indiana.
\item We held out 5\% of the trees from all of the trees in the dataset
for the midwestern subdomain (leaving aside the held-out Minnesota
subregion cells). This was meant to assess the ability of the model
to estimate the composition in cells in which data were available. 
\end{enumerate}
Finally, in a separate sensitivity analysis we instead left out 80\%
of the cells in Minnesota subregion at random. This variation on the
first experiment above was meant to indicate whether our model comparison
conclusions would be robust as the digitization process for Illinois
and Indiana progresses and provides us with increasingly dense data.

@@

Following the suggestions in \cite{Gnei:etal:2007}, we considered
the following metrics: Brier score, log predictive density, mean square
prediction error, mean absolute error, and coverage and length of
prediction intervals. Further details on each are given below. For
experiment 1, we define $Y_{i}=\{Y_{i1},\ldots,Y_{iP}\}$ as the count
of all trees in held-out cell \emph{i} and for experiment 2, $Y_{i}$
is the count of held-out individual trees in the cell, while $y_{ijp}$
is an indicator variable taking value either 0 or 1 depending on whether
the \emph{j}th held-out tree in the \emph{i}th cell is of taxon \emph{p}. $\hat{\theta}_{ip}=Y_{ip}/n_{i}$
is the empirical proportion in category \emph{p} for the $n_{i}$ held-out
trees in cell \emph{i}. We calculated each of the metrics in two ways.
First, we used the posterior mean composition estimates (as an evaluation
of our core predictions), with $\tilde{\theta}_{p}(s)$ being the
posterior mean. Second, we averaged the metric over the posterior
samples (as an evaluation of our full data product, including uncertainty),
taking $\tilde{\theta}_{p}(s)$ to be an individual MCMC sample and
then averaging the metric over all the posterior samples. 
\begin{enumerate}
\item Brier score: \cite{Gnei:etal:2007} suggest this metric, which has
been in use for decades. For multi-category as opposed to binary outcomes,
this takes the form
\[
\frac{1}{n}\sum_{i=1}^{m}\sum_{j=1}^{n_{i}}\sum_{p=1}^{P}(y_{ijp}-\tilde{\theta}_{p}(s_{i}))^{2}
\]
where $n=\sum_{i=1}^{m}n_{i}$ is the total number of held-out trees
for a given experiment and \emph{j} indexes across held-out trees in cell
\emph{i}. 
\item Log predictive density: This metric takes the log of the probability
density of held-out observations under the fitted model, $Y_{i}\sim\mbox{Multinom}(n_{i},\{\tilde{\theta}_{1}(s_{i}),\ldots,\tilde{\theta}_{P}(s_{i})\})$,
summing on the log scale across all of the held-out data.

@@

\item (Experiment 1 only) Weighted root mean square prediction error (RMSPE),

\[
\sqrt{\frac{1}{Pn}\sum_{i=1}^{m}\sum_{p=1}^{P}n_{i}(\hat{\theta}_{ip}-\tilde{\theta}_{p}(s))^{2}}
\]
and mean absolute error (MAE)
\[
\frac{1}{Pn}\sum_{i=1}^{m}\sum_{p=1}^{P}n_{i}|\hat{\theta}_{ip}-\tilde{\theta}_{p}(s)|:
\]
These metrics calculate the error of the estimated proportions relative
to the empirical proportions based on the held-out trees, averaging
over cells and taxa. We weight by the number of held-out trees in
each cell to account for the greater variability in the empirical
proportions in locations with few held-out trees. 
\item (Experiment 1 only) Coverage and length of 95\% prediction intervals
for $Y_{ip}$. We considered only cells with at least 50 trees to
focus our assessment on cases where empirical proportions were reasonably
certain and avoid being strongly influenced by predictive inference
for cells where observational variability dominates.
\end{enumerate}
Note that all of the metrics except coverage and interval length can
be applied to individual posterior samples and therefore allow us
to estimate the posterior probability that one model has a lower (better)
value of the metric than the other model by simply calculating the
proportion of samples for which the model has a lower value of the
metric. Also note that in addition to allowing comparison
between models the MAE and RMSPE metrics allow one to assess absolute
performance of each model in predicting composition.

@@

Here we summarize the results of our analyses that inform the choice
between the CAR and SPDE models.

@@

For this final data product, we ran the model using the CAR specification
with all of the data (including the data held out
in the model comparison analyses) for 150,000 iterations with the
same burn-in and subsampling details as described in Section \ref{sec:Model-comparison}.
Based on graphical checks and calculation of effective sample size
values, mixing was generally reasonable, but for some of the hyperparameters
was relatively slow, particularly for less common taxa. Despite this,
mixing for the variables of substantive interest -- the proportions
-- was good, with effective sample sizes for the final product generally
near 250.

>>>>

In \emph{Data} we describe the data sources, while in \emph{Statistical model} we describe 
the statistical model used to create the data product. In In \emph{Model comparison}
 we quantitatively compare competing
statistical specifications, and in \emph{Data Product}
we describe the final data product. In \emph{Discussion}
we discuss \textcolor{red}{the uncertainties estimated by }and the
limitations of the statistical model, and we list related data products
under development.

@@

The raw data were obtained from land division survey records collated
and digitized from across the northeastern U.S. by a number of researchers
(Fig. \ref{fig:Spatial-domain}). For the states of Minnesota, Wisconsin,
Illinois, Indiana, and Michigan (the midwestern subdomain), digitized
data are available at PLS survey point locations and have been aggregated
to a regular 8 km grid in the Albers projection. (Note that for Indiana
and Illinois, at the moment trees are associated with township centroids
and then assigned to 8 km grid cells based on the centroid, but in
the near future we will have point locations available for each tree.)
For the states of Ohio, Pennsylvania, New Jersey, New York and the
six New England states (the eastern subdomain), data are aggregated
at the township level. \textcolor{red}{We make predictions for all
of the states listed above; these constitute our core domain. }There
are also data from a single township in Quebec and a single township
in northern Delaware\textcolor{red}{; these data help inform predictions
in nearby locations within our core domain, but predictions are not
made for Quebec or Delaware}. Digitization of PLS data in Minnesota,
Wisconsin and Michigan is essentially complete, with PLS data for
nearly all 8 km grid cells, but data in Illinois and Indiana represent
a sample of the full set of grid cells, with survey record transcription
ongoing. Data for the eastern states are available for a subset of
the full set of townships covering the domain; the TPS data for some
townships were lost, incomplete, or have not been located \cite{Cogb:etal:2002}.

@@

Extensive details on the upper Midwest (Minnesota, Wisconsin, Michigan)
data and processing steps are available \cite{goring2015composition};
key elements include the use of only corner points, the use of only
the two closest trees at each corner point, spatially-varying correction
factors for sampling effort, and a standardized taxonomy table. The
lower Midwest (Illinois, Indiana) data were purchased from the Indiana
State Archives (Indiana) and Hubtack Document Resources (hubtack.com;
Illinois) and processed using similar steps as for the upper Midwest
data. Digitization of the Illinois and Indiana data is still underway,
so many grid cells contained no data at the time the statistical model
was fit. Note that the number of trees per grid cell varies depending
on the number of survey points in a cell, with an average of 124 trees
per cell. The gridded data at the 8 km resolution for the midwest
subdomain are available through the NIS data portal \textcolor{red}{as
DOI:10.5072/FK2/e9c144ab83af70e29b8d6647aaae58f5} \textcolor{red}{\cite{Gori:etal:data:2016}}.
The TPS data were compiled by C.V. Cogbill from a myriad of archival
sources representing land division surveys conducted in connection
with local settlement\textcolor{red}{{} and are available through the
NIS data portal as DOI:10.5072/FK2/4d44aebe71dbf65290d8ce1bde03598b
and DOI:10.5072/FK2/88e167487b22d5f657e7ce7e033772e1 \cite{Cogb:dataNE:2016,Cogb:dataOH:2016}}.

@@

Our domain is a rectangle covering all of the states using a metric
Albers (Great Lakes and St. Lawrence) projection (PROJ4: EPSG:3175),
with the rectangle split into 8 km cells, arranged in a 296 by 180
grid of cells, with the centroid of the cell in the southwest corner
located at (-71000 m, 58000 m). For the midwestern subdomain we use
the western-most 146 by 180 grid of cells\textcolor{red}{{} when fitting
the statistical models}. For the eastern subdomain we use the eastern-most
180 by 180 grid of cells and then omit 23 rows of cells in the north
and 17 rows of cells in the south, as these grid cells are outside
of the states containing data.

@@

We fit a Bayesian statistical model to the data, with two primary
goals:
\begin{enumerate}
\item To estimate composition on a regular grid across the entire domain,
filling gaps where no data are available, and
\item To quantify uncertainty in composition at all locations. Even in grid
cells and townships with data, we wish to quantify uncertainty because
the empirical proportions represent estimates of the true proportions
that could be calculated using the full population of all the trees
in a grid cell or township.
\end{enumerate}
\textcolor{red}{At a high level, the Bayesian statistical model estimates
composition across the domain, even in locations with sparse or no
data, by combining the raw composition data with the assumption that
composition varies in a smooth spatial fashion across the domain.
The information in the data is quantified by the data model, also
known as the likelihood. The assumption of smoothness is built into
the model by representing the true unknown spatially-varying composition
using a statistical spatial process representation that induces smoothing
of estimates across nearby locations. This spatial process representation
is a form of prior distribution and is a function of model parameters
called hyperparameters that determine the correlation structure of
the process and are also estimated based on the data. }

@@

\textcolor{red}{In the remainder of this section we provide the technical
specification of the model and of the computations involved in fitting
the model.}

@@

The proportions, $\theta_{p}(s_{i}),\, p=1,\ldots,P$, are modeled
spatially by a set of \emph{P} Gaussian spatial processes, one per taxon,
$\alpha_{p}(s_{i}),\, p=1,\ldots,P$. This collection of processes
defines a multivariate spatial process for composition. The $\alpha_{p}(s)$
processes are defined on the 8 km grid, $\alpha_{p}=\{\alpha_{p}(s_{1}),\ldots,\alpha_{p}(s_{m})\}$
for the \emph{m} grid cells. In  \emph{Latent variable model}
we introduce a multinomial probit model that relates the $\alpha_{p}(s)$
processes to the proportion processes, $\theta_{p}(s)$, via the introduction
of latent variables, with an implicit sum-to-one constraint, $\sum_{p=1}^{P}\theta_{p}(s)=1$.
\textcolor{red}{A multinomial probit model is similar to logistic
regression, used for modeling a binary outcome based on an underlying
probability the outcome will occur, but generalizes to modeling a
categorical variable based on probabilities for each category. }

The critical component of the statistical model is the representation
of $\alpha_{p}(s)$ as a spatial process. \textcolor{red}{A spatial
process is a statistical representation that models spatially-correlated
values. It provides} a prior structure that serves to smooth across
noise in the observations and allows for prediction at locations based
on information from nearby locations, including interpolation to locations
with no data. Apart from the sum-to-one constraint, the taxa are considered
to be independent in the prior. We did not want to impose any structure
that ties the different taxa together, as any correlation will likely
vary across space.

In the next section, we consider two spatial models to define the
structure of the $\alpha_{p}(s)$ processes, a standard conditional
autoregressive model \cite{Bane:etal:2004} and a Gaussian Markov
random field (MRF) approximation to a Gaussian process with \matern
covariance \cite{Lind:etal:2011}. \textcolor{red}{These models are
specific statistical formulations of spatial processes that represent
spatial correlation by defining neighborhoods around each location
that are used to help inform predictions at the location. }

@@

MRF models \textcolor{red}{represent the neighborhood information
}by working directly with the precision matrix \textcolor{red}{(the
inverse of the covariance matrix)} of the values of the spatial process,
so calculation of the prior density of $\alpha_{p}$ is computationally
simple \cite{Rue:Held:2005}. \textcolor{red}{However,} in situations
\textcolor{red}{in which} the likelihood is not normal, \textcolor{red}{such
as our multinomial likelihood}, it can be difficult to set up effective
MCMC algorithms that are able to move in the high-dimensional space
of $\alpha_{p}$. The latent variable representation
helps to alleviate this problem. Next we describe two alternative
spatial models that we considered; in \emph{Model comparison},
\textcolor{red}{we evaluate the models on held-out data} to choose
between the two.

@@

Our first model is a standard conditional autoregressive (CAR) model\textcolor{red}{;
technical details can be found in} \cite{Bane:etal:2004}. We use
a standard form of this model \textcolor{red}{that} treats the four
cardinal neighbors of each grid cell as the neighbors of the grid
cell. The corresponding precision matrix has diagonal elements, $Q_{ii}$,
equal to the number of neighbors for the \emph{i}th area (i.e., four except
for cells on the boundary of the domain), while $Q_{ik}=-1$ (the
negative of a weight of one) when areas \emph{i} and \emph{k} are neighbors
and $Q_{ik}=0$ when they are not. This gives the following model
for the values of $\alpha_{p}(s_{i})$ collected as a vector across
all of the grid cells, $i=1,\ldots,m$: 
\[
\alpha_{p}\sim\mbox{N}(0,\sigma_{p}^{2}Q^{-}).
\]
The use of the generalized inverse notation indicates that \emph{Q} is
not full-rank, but is of rank $m-1$; this gives an improper prior
on an implicit overall mean for the process values. \textcolor{red}{Note
that we specify an explicit mean of zero because a non-zero mean would
not be identifiable in light of the implicit mean.} This specification
is called an \textit{intrinsic conditional autoregression (ICAR)}
and we can write $Q=D-C$ where \emph{C} is the $m\times m$ adjacency
matrix defining the neighborhood relation of the locations; that is
$C{}_{ik}=1$ if locations \emph{i} and \emph{k} are neighbors and zero otherwise.
The matrix $D$ is an $m\times m$ diagonal matrix containing the
row sums of matrix $C$ as the diagonal entries, $D{}_{ii}={\displaystyle \sum_{k=1}^{m}C{}_{ik}}.$

@@

Gaussian processes (GP) are also standard models for spatial processes \cite{Bane:etal:2004}.
GP models are computationally challenging for large datasets because
of \textcolor{red}{computational} manipulations involving large covariance
matrices. Given this, \cite{Lind:etal:2011} proposed a new framework
for using Gaussian MRFs (GMRFs) as approximations to GPs, based on
the use of stochastic partial differential equations (SPDEs).

\textcolor{red}{Gaussian processes are generally constructed using
one of a number of correlation functions that define how the strength
of correlation between the values of the process at two locations
decays as a function of the distance between the locations. }We consider
Gaussian processes in the \textcolor{red}{commonly-used} \matern
class, using the following parameterization of the \matern correlation
function, 
\begin{equation}
R(d)=\frac{1}{\Gamma(\nu)2^{\nu-1}}\left(\frac{2\sqrt{\nu}d}{\rho}\right)^{\nu}\mathcal{K}_{\nu}\left(\frac{2\sqrt{\nu}d}{\rho}\right),\label{eq:Matern}
\end{equation}
where $d$ is Euclidean distance, $\rho$ is the spatial range parameter,
and $\mathcal{K}_{\nu}(\cdot)$ is the modified Bessel function of
the second kind, whose order is the smoothness (differentiability)
parameter, $\nu>0$. $\nu=0.5$ gives the exponential covariance.
For any pair of locations, $R(d)$ defines the correlation of the
process, (i.e., $\alpha_{p}(s)$ in our context), as a function of
the distance between the locations. Considering all pairs of locations,
this defines a correlation matrix for all locations of interest.

The approach of \cite{Lind:etal:2011} allows us to consider MRF
approximations to the \matern-based GP for $\nu=1$ and $\nu=2$.
Our second spatial model is this Lindgren approximation for \matern-based
GPs with $\nu=1$. To implement this Lindgren model, one modifies
the \emph{Q} matrix defined previously as follows \textcolor{red}{based
on the technical specification of the precision matrix provided in
\cite{Lind:etal:2011}}. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
elements of \emph{Q} are $4+a^{2}$. The entries corresponding to cardinal
neighbors are $-2a$. Those for diagonal neighbors are $2$, and those
for 2nd-order cardinal neighbors are $1$. This extends the neighborhood
structure relative to the CAR model and parameterizes it as a function
of $\rho$.

The primary difference between the CAR and Lindgren models is that
the Lindgren model provides an additional degree of freedom by estimating
$\rho$. In particular $\rho$ allows us to estimate the locality
of the \textcolor{red}{spatial} smoothing. As $\rho$ decreases, the
model uses increasingly localized data to estimate the compositional
proportions at a given location, effectively averaging the empirical
proportions over smaller neighborhoods. In general, the \cite{Lind:etal:2011}
model will generally provide for a smoother estimate than the CAR
model \cite{Paci:2013}.

@@

\noindent The ICAR specification contains a set of hyperparameters
$\{\sigma_{p}^{2}\},$ $p=1,\ldots P$. Following \cite{Gelm:2006}
we use a uniform distribution on each $\sigma_{p}$ parameter, with
upper bound of 1000. For the SPDE model we also have hyperparameters
$\{\mu_{p}\}$, which we give flat, non-informative priors (truncated
at $\pm10$), and $\{\rho_{p}\}$, which we give uniform priors on
the interval $(0.1,\exp(5))$. \textcolor{red}{These various hyperparameters
are unknown parameters that control the spatial structure of the two
spatial models and are estimated from both the data and the prior
distributions just specified based on the Bayesian approach.}

@@

We compared the CAR and SPDE models \textcolor{red}{by holding out
data} from the fitting process and assessing the fit of the model
on the held-out data. We used two experiments\textcolor{red}{{} with
held-out data}:
\begin{enumerate}
\item The first experiment used a subregion containing most of Minnesota
and a small amount of western Wisconsin, defined to be the cells whose
x-coordinate was less than 300,000 m (this defines a north-south line
that approximately goes through Duluth, Minnesota) and hereafter referred
to as the ``Minnesota subregion''. We chose this subregion for \textcolor{red}{evaluation}
because of its high data density, allowing us to experiment with the
effects of increasing data sparsity on model performance. We held
out all data from 95\% of the cells in this Minnesota subregion, with
cells selected at random. This was meant to assess the ability of
the model to interpolate from a sparse set of cells/townships and
mimics the limited data in Illinois and Indiana.
\item We held out 5\% of the trees from all of the trees in the dataset
for the midwestern subdomain (leaving aside the held-out Minnesota
subregion cells). This was meant to assess the ability of the model
to estimate the composition in cells in which data were available. 
\end{enumerate}
Finally, in a separate sensitivity analysis we instead left out 80\%
of the cells in Minnesota subregion at random. This variation on the
first experiment above was meant to indicate whether our model comparison
conclusions would be robust as the digitization process for Illinois
and Indiana progresses and provides us with increasingly dense data.

@@

Following the suggestions in \cite{Gnei:etal:2007}, we considered
the following metrics: Brier score, log predictive density, mean square
prediction error, mean absolute error, and coverage and length of
prediction intervals. Further details on each are given below. For
experiment 1, we define $Y_{i}=\{Y_{i1},\ldots,Y_{iP}\}$ as the count
of all trees in held-out cell \emph{i} and for experiment 2, $Y_{i}$
is the count of held-out individual trees in the cell, while $y_{ijp}$
is an indicator variable taking value either 0 or 1 depending on whether
the \emph{j}th held-out tree in the \emph{i}th cell is of taxon \emph{p}. \textcolor{red}{$\hat{\theta}_{ip}=Y_{ip}/n_{i}$
is the empirical proportion in category \emph{p} for the $n_{i}$ held-out
trees in cell \emph{i}.} We calculated each of the metrics in two ways.
First, we used the posterior mean composition estimates (as an evaluation
of our core predictions), with $\tilde{\theta}_{p}(s)$ being the
posterior mean. Second, we averaged the metric over the posterior
samples (as an evaluation of our full data product, including uncertainty),
taking $\tilde{\theta}_{p}(s)$ to be an individual MCMC sample and
then averaging the metric over all the posterior samples. 
\begin{enumerate}
\item Brier score: \cite{Gnei:etal:2007} suggest this metric, which has
been in use for decades. For multi-category as opposed to binary outcomes,
this takes the form
\[
\frac{1}{n}\sum_{i=1}^{m}\sum_{j=1}^{n_{i}}\sum_{p=1}^{P}(y_{ijp}-\tilde{\theta}_{p}(s_{i}))^{2}
\]
where $n=\sum_{i=1}^{m}n_{i}$ is the total number of held-out trees
for a given experiment and \emph{j} indexes across held-out trees in cell
\emph{i}. 
\item Log predictive density: This metric takes the log of the \textcolor{red}{probability
}density of held-out observations under the fitted model, $Y_{i}\sim\mbox{Multinom}(n_{i},\{\tilde{\theta}_{1}(s_{i}),\ldots,\tilde{\theta}_{P}(s_{i})\})$,
summing on the log scale across all of the held-out data.

@@

\item (Experiment 1 only) Weighted root mean square prediction error (RMSPE),
\textcolor{red}{
\[
\sqrt{\frac{1}{Pn}\sum_{i=1}^{m}\sum_{p=1}^{P}n_{i}(\hat{\theta}_{ip}-\tilde{\theta}_{p}(s))^{2}}
\]
}and mean absolute error (MAE)\textcolor{red}{
\[
\frac{1}{Pn}\sum_{i=1}^{m}\sum_{p=1}^{P}n_{i}|\hat{\theta}_{ip}-\tilde{\theta}_{p}(s)|:
\]
}These metrics calculate the error of the estimated proportions relative
to the empirical proportions based on the held-out trees, averaging
over cells and taxa. We weight by the number of held-out trees in
each cell to account for the greater variability in the empirical
proportions in locations with few held-out trees. 
\item (Experiment 1 only) Coverage and length of 95\% prediction intervals
for $Y_{ip}$. We considered only cells with at least 50 trees to
focus our assessment on cases where empirical proportions were reasonably
certain and avoid being strongly influenced by predictive inference
for cells where observational variability dominates.
\end{enumerate}
Note that all of the metrics except coverage and interval length can
be applied to individual posterior samples and therefore allow us
to estimate the posterior probability that one model has a lower (better)
value of the metric than the other model by simply calculating the
proportion of samples for which the model has a lower value of the
metric. \textcolor{red}{Also note that in addition to allowing comparison
between models the MAE and RMSPE metrics allow one to assess absolute
performance of each model in predicting composition.}

@@

Here we summarize the results of our \textcolor{red}{analyses} that inform the choice
between the CAR and SPDE models.

@@

For this final data product, we ran the model using the CAR specification
\textcolor{red}{with all of the data (including the data held out
in the model comparison analyses)} for 150,000 iterations with the
same burn-in and subsampling details as described in Section \ref{sec:Model-comparison}.
Based on graphical checks and calculation of effective sample size
values, mixing was generally reasonable, but for some of the hyperparameters
was relatively slow, particularly for less common taxa. Despite this,
mixing for the variables of substantive interest -- the proportions
-- was good, with effective sample sizes for the final product generally
near 250.
